---
title: "Figure 6D code"
author: "Anna Kogan"
date: "2026-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(dplyr)
library("RColorBrewer")
library(scales)
```

"Main folder" - folder on the drive with the whole experiment In this main folder 1 sub folder = 1 experimental sample "All directories" - a full list (with paths) of the directories in the folder

DAPI-masked data

```{r TS_Folder}
main_folder <- "E:/M096_DAPI"
all_dirs <- paste(list.files(main_folder, full.names = TRUE), "/", sep = "")
all_dir_names <- list.files(main_folder, full.names = FALSE)
print(paste("Number of folders:",length(all_dirs), sep = " "))
print(paste("The first few folders:", paste(all_dir_names[1], all_dir_names[2], all_dir_names[3], all_dir_names[4], sep = " "), sep = " "))
```
```{r Column_names}
subfolder.names <- list.files(main_folder, full.names = TRUE)

#collects subfolder names for the sample names, change dependent on your pattern
colnames.all.files <- list.files(main_folder)

for (i in 1:length(colnames.all.files)){
  colnames.all.files[i] <-
    sub("_", " ", substr(colnames.all.files[i], 6, nchar(colnames.all.files[i])))
}

#collects subfolder names for the sample names, input your pattern
colnames.merged.files <- list.files(main_folder)[(1:2)*3-2]
for (i in 1:length(colnames.merged.files)){
  colnames.merged.files[i] <-
    sub("_", " ", substr(colnames.merged.files[i], 6, nchar(colnames.merged.files[i])-2))
}

```


Analysis for nascent transcription - counts number of transcription sites per cell:

Collect two databases: one with individual replicates, one with merged replicates

```{r TS_count_all}

#this is list of file names
arrays3D <- list.files(subfolder.names, pattern = "700_25_count_In_Result",full.names = TRUE)

first.file <- read.delim(arrays3D[1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(first.file) <- c("cell", "1")

#cbinds the other files using first column to merge them
for(i in 2:(length(arrays3D))){

  file1 <- read.delim(arrays3D[i], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file1) <- c("cell", (i))
  
  first.file <- merge(first.file, file1, 
                      by = "cell", all = TRUE)
} 

#remove first column, apply gather, omit NAs
TS_count.all <- first.file[,-1]
#write.csv(x = TS_count.all, file = paste(main_folder, "/TS_count_all_Result.csv", sep=""))


colnames(TS_count.all) <- colnames.all.files

TS_count.all_gathered <- na.omit(gather(TS_count.all, key = "cell", value = "TS_count"))

TS_count.all_gathered$ones <- 1
#order the data and some other data manipulations 
TS_count.all_gathered <- TS_count.all_gathered[order(TS_count.all_gathered$cell, TS_count.all_gathered$TS_count),]

#calculating frequency of specific numbers of TS 
options(dplyr.summarise.inform = FALSE)
TS_count.all_gathered <- TS_count.all_gathered %>% group_by(cell, TS_count) %>% summarise(Freq= n())

```

```{r TS_count_merged}

arrays3D <- list.files(subfolder.names, pattern = "700_25_count_In_Result",full.names = TRUE)

first.file <- read.delim(arrays3D[1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(first.file) <- c("cell", "1")

second.file <- read.delim(arrays3D[2], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(second.file) <- c("cell", "1")

third.file <- read.delim(arrays3D[3], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(third.file) <- c("cell", "1")

first.file <- rbind(first.file, second.file, third.file)

first.file[,1] <- 1:nrow(first.file)

#cbinds the other files using first column to merge them
for(i in 2:(length(arrays3D)/3)){
  file1 <- ""
  file2 <- ""
  file3 <- ""
  
  file1 <- read.delim(arrays3D[3*i-2], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file1) <- c("cell", (3*i-2))
  
  file2 <- read.delim(arrays3D[3*i-1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file2) <- c("cell", (3*i-2))
  
  file3 <- read.delim(arrays3D[3*i], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file3) <- c("cell", (3*i-2))
  
  file1 <- rbind(file1, file2, file3)
  
  file1[,1] <- 1:nrow(file1)
  first.file <- merge(first.file, file1, 
                      by = "cell", all = TRUE)
}

#remove first column, apply gather, omit NAs
TS_count.merged <- first.file[,-1]
colnames(TS_count.merged) <- colnames.merged.files

TS_count.merged_gathered <- na.omit(gather(TS_count.merged, key = "cell", value = "TS_count"))

TS_count.merged_gathered$ones <- 1
#order the data and some other data manipulations 
TS_count.merged_gathered <- TS_count.merged_gathered[order(TS_count.merged_gathered$cell, TS_count.merged_gathered$TS_count),]

#calculating frequency of specific numbers of TS 
options(dplyr.summarise.inform = FALSE)
TS_count.merged_gathered <- TS_count.merged_gathered %>% group_by(cell, TS_count) %>% summarise(Freq= n())

```

Plots:

```{r TS_count_all_plot}
ggplot(TS_count.all_gathered, aes(fill = TS_count, y = Freq, x = cell))+
  scale_fill_gradient2(limits=c(0, 12), low=muted("dark red"), mid="floral white", high="light blue", midpoint=6)+
  geom_bar(position = "fill", stat = "identity")+
  labs(x ="", y = "Proportion of cells")+
  theme_pubr()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

#for analysis in Prism:
#TS_count_stats <- sapply(TS_count.all, mean, na.rm = TRUE)

```

```{r TS_count_merged_plot}
ggplot(TS_count.merged_gathered, aes(fill = TS_count, y = Freq, x = cell))+
  scale_fill_gradient2(limits=c(0, 12), low=muted("dark red"), mid="floral white", high="light blue", midpoint=6)+
  geom_bar(position = "fill", stat = "identity")+
  labs(x ="", y = "Proportion of cells")+
  theme_pubr()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
