---
title: "Figure 4E code"
author: "Anna Kogan"
date: "2026-09-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(dplyr)
library("RColorBrewer")
library(scales)
```

"Main folder" - folder on the drive with the whole experiment In this main folder 1 sub folder = 1 experimental sample "All directories" - a full list (with paths) of the directories in the folder

```{r Folder}
main_folder <- "D:/Anna/M037_9_GBP5/"
all_dirs <- paste(list.files(main_folder, full.names = TRUE), "/", sep = "")
all_dir_names <- list.files(main_folder, full.names = FALSE)

print(paste("Number of folders:",length(all_dirs), sep = " "))
print(paste("The first few folders:", paste(all_dir_names[1], all_dir_names[2], all_dir_names[3], all_dir_names[4], sep = " "), sep = " "))
```
This chunk of the code goes through each folder and finds the results from the 3D object counter (csv files) 1) Generates a csv with mRNA counts per cell for each sample 2) Generates a csv with the number of TS per cell for each sample 3) Generates a csv with a list of all TS and their integrated densities for each sample

Commented out - bits that could generate a histogram of each of those csv and save it in the same folder, run if needed

NB! Modify the intensity and size threshold that were used in 3D object counter to correctly detect and label files

```{r Make_SCV_files_for_each_sample}

for (d in 1:length(all_dirs)) {

dir_masks <- paste(all_dirs[d], "OutputMASK/masks/singlecellmasks/", sep = "")
subfolder.names<-list.dirs(dir_masks)

#### mRNA ####
#collects count of mRNA per cell to the out.file
out.file<-"" 
arrays3D <- list.files(subfolder.names, pattern = "160_10",full.names = TRUE)
for(i in 1:length(arrays3D)){
  new_file <- read.delim(arrays3D[i], header = TRUE, sep = ",", dec = ".")
  spot_count <- nrow(new_file)
  out.file <- rbind(out.file, spot_count)
}

out.file <- out.file[-1]
out.file <- as.integer(out.file)

write.csv(x = out.file, file = paste(all_dirs[d], all_dir_names[d], "_160_10_Ex_Result.csv", sep=""))

#### mRNA plot ####
# mRNA_count <- data.frame(out.file)
# mRNA <- ggplot(mRNA_count, aes(x=out.file)) +
#   stat_count()+
#   labs(y = "", x = "Number of mRNA per cell") +
#   theme_pubr()
# mRNA
# ggsave(plot = mRNA, device = "pdf", main_folder = all_dirs[d], filename = paste(all_dir_names[d],"_mRNA.pdf"))

}

```

```{r Column_names}
subfolder.names <- list.files(main_folder, full.names = TRUE)

#collects subfolder names for the sample names, change dependent on your pattern
colnames.all.files <- list.files(main_folder)

for (i in 1:length(colnames.all.files)){
  colnames.all.files[i] <-
    sub("_", " ", sub("_", " ", colnames.all.files[i]))
}

#collects subfolder names for the sample names, input your pattern
colnames.merged.files <- list.files(main_folder)[(1:16)*3-2]
for (i in 1:length(colnames.merged.files)){
  colnames.merged.files[i] <-
    sub("_", " ", sub("_", " ", substr(colnames.merged.files[i], 1, nchar(colnames.merged.files[i])-2)))
}


print(paste("The first few replicate names:",paste(colnames.all.files[1:5], collapse = ', '), sep = " "))

print(paste("The first few sample names:",paste(colnames.merged.files[1:3], collapse = ', '), sep = " "))



```

Analysis for mRNA:

Collect two databases: one with individual replicates, one with merged replicates

```{r mRNA_all}

#makes a list of csv files:
arrays3D <- list.files(subfolder.names, pattern = "_160_10_Ex_Result",full.names = TRUE)

#reads the first file
first.file <- read.delim(arrays3D[1], header = TRUE, sep = ",", dec = ".")
colnames(first.file) <- c("cell", "1")

#cbinds the other files using first column to merge them
for(i in 2:(length(arrays3D))){
  new_file <- ""
  new_file <- read.delim(arrays3D[i], header = TRUE, sep = ",", dec = ".") 
  colnames(new_file) <- c("cell", (i))
  first.file <- merge(first.file, new_file, 
                    by = "cell", all = TRUE)
  
}

mRNA_all  <-  first.file[,-1]
colnames(mRNA_all) <- colnames.all.files

# ##for further Prism analysis:
# mRNA_means <- data.frame(
#   Column_Names = colnames(mRNA_all),
#   Ave = sapply(mRNA_all, mean, na.rm = TRUE)
# )
# mRNA_means$Ave
# 
# 
# write.csv(mRNA_all, file = "D:/Anna/M037_9_mRNA_all.csv")

```

```{r mRNA_merged}

#makes a list of csv files:
arrays3D <- list.files(subfolder.names, pattern = "_160_10_Ex_Result",full.names = TRUE)

#reads the first 3 files
first.file <- read.delim(arrays3D[1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(first.file) <- c("cell", "1")

second.file <- read.delim(arrays3D[2], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(second.file) <- c("cell", "1")

third.file <- read.delim(arrays3D[3], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(third.file) <- c("cell", "1")

first.file <- rbind(first.file, second.file, third.file)
first.file[,1] <- 1:nrow(first.file)


#cbinds the other files using first column to merge them
for(i in 2:(length(arrays3D)/3)){
  
  file1 <- read.delim(arrays3D[3*i-2], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file1) <- c("cell", (3*i-2))
  
  file2 <- read.delim(arrays3D[3*i-1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file2) <- c("cell", (3*i-2))
  
  file3 <- read.delim(arrays3D[3*i], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file3) <- c("cell", (3*i-2))
  
  file1 <- rbind(file1, file2, file3)
  
  file1[,1] <- 1:nrow(file1)
  first.file <- merge(first.file, file1, 
                      by = "cell", all = TRUE)
  
}

mRNA_merged  <-  first.file[,-1]
colnames(mRNA_merged) <- colnames.merged.files
```

Plot:

```{r mRNA_all_plot}
#prepare for plotting
mRNA_all_gathered <- na.omit(gather(mRNA_all, key="sample",value = "mRNA_per_cell" ))
#plot
ggplot(mRNA_all_gathered, aes(y = mRNA_per_cell, x = sample)) +
  geom_violin(trim = TRUE, fill="peru", scale = "width",
             draw_quantiles = c(0.25, 0.5, 0.75)) +
  stat_summary_bin(fun = "mean", geom = "bar", colour = "black", fill = "white", alpha = 0.5)+
  labs(title="Distribution of mRNA",
       x ="", y = "mRNA per cell")+
  theme_pubr() +
  theme(axis.text.x=element_text(angle = 60, hjust = 1))
```

Plot used in Figure 4E:

```{r mRNA_merged_plot}
#prepare for plotting
mRNA_merged_gathered <- na.omit(gather(mRNA_merged, key="sample",value = "mRNA_per_cell" ))
#plot
ggplot(mRNA_merged_gathered, aes(y = mRNA_per_cell, x = sample)) +
  geom_violin(trim = TRUE, fill="peru", scale = "width",
             draw_quantiles = c(0.25, 0.5, 0.75)) +
  stat_summary_bin(fun = "mean", geom = "bar", colour = "black", fill = "white", alpha = 0.5)+
  labs(title="Distribution of mRNA",
       x ="", y = "mRNA per cell")+
  theme_pubr() +
  theme(axis.text.x=element_text(angle = 60, hjust = 1))
```
