---
title: "Figures S6C S8D"
author: "Anna Kogan"
date: "2026-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Libraries, message=FALSE, warning=FALSE}
library(ggplot2)
library(ggpubr)
library(tidyverse)
library(dplyr)
library("RColorBrewer")
library(scales)
```

"Main folder" - folder on the drive with the whole experiment In this main folder 1 sub folder = 1 experimental sample "All directories" - a full list (with paths) of the directories in the folder

```{r Folder}
main_folder <- "E:/M040_tiffs/"
all_dirs <- paste(list.files(main_folder, full.names = TRUE)[1:18], "/", sep = "")
all_dir_names <- list.files(main_folder, full.names = FALSE)[1:18]
print(paste("Number of folders:",length(all_dirs), sep = " "))
print(paste("The first few folders:", paste(all_dir_names[1], all_dir_names[2], all_dir_names[3], all_dir_names[4], sep = " "), sep = " "))
```

This chunk of the code goes through each folder and finds the results from the 3D object counter (csv files) 1) Generates a csv with mRNA counts per cell for each sample 2) Generates a csv with the number of TS per cell for each sample 3) Generates a csv with a list of all TS and their integrated densities for each sample

Commented out - bits that could generate a histogram of each of those csv and save it in the same folder, run if needed

NB! Modify the intensity and size threshold that were used in 3D object counter to correctly detect and label files

```{r Make_SCV_files_for_each_sample}

for (d in 1:length(all_dirs)) {

dir_masks <- paste(all_dirs[d], "OutputMASK/masks/singlecellmasks/", sep = "")
subfolder.names<-list.dirs(dir_masks)


#### mRNA ####
#collects count of mRNA per cell to the out.file
out.file<-""
arrays3D <- list.files(subfolder.names, pattern = "160_10",full.names = TRUE)
for(i in 1:length(arrays3D)){
  new_file <- read.delim(arrays3D[i], header = TRUE, sep = ",", dec = ".")
  spot_count <- nrow(new_file)
  out.file <- rbind(out.file, spot_count)
}

out.file <- out.file[-1]
out.file <- as.integer(out.file)

write.csv(x = out.file, file = paste(all_dirs[d], all_dir_names[d], "_160_10_Ex_Result.csv", sep=""))

# ### mRNA plot ####
# mRNA_count <- data.frame(out.file)
# mRNA <- ggplot(mRNA_count, aes(x=out.file)) +
#   stat_count()+
#   labs(y = "", x = "Number of mRNA per cell") +
#   theme_pubr()
# mRNA
# ggsave(plot = mRNA, device = "pdf", path = all_dirs[d], filename = paste(all_dir_names[d],"_mRNA.pdf", sep = ""))

#### TS count ####
#collects count of TS per cell to the out.file
arrays3D <- list.files(subfolder.names, pattern = "500_25",full.names = TRUE)
out.file<-""
for(i in 1:length(arrays3D)){
  new_file <- read.delim(arrays3D[i], header = TRUE, sep = ",", dec = ".")
  new_file <- new_file[new_file$Volume..px.3.>100,]
  spot_count <- nrow(new_file)
  out.file <- rbind(out.file, spot_count)
}

out.file <- out.file[-1]
out.file <- as.integer(out.file)
write.csv(x = out.file, file = paste(all_dirs[d], all_dir_names[d], "_500_25_Vol_count_In_Result.csv", sep=""))

# #### TS count plot ####
# TS_count <- data.frame(out.file) #saves TS per cell to TS_count
# 
# NofTS <- ggplot(TS_count, aes(x=out.file)) +
#   stat_count()+
#   labs(y = "", x = "Number of TS per cell") +
#   theme_pubr()
# NofTS #could save this
# ggsave(plot = NofTS, path = all_dirs[d], device = "pdf", filename = paste(all_dir_names[d],"_N_of_TS.pdf"))

}

```

```{r Column_names}
subfolder.names <- list.files(main_folder, full.names = TRUE)

#collects subfolder names for the sample names, change dependent on your pattern
colnames.all.files <- list.files(main_folder)[1:18]

for (i in 1:length(colnames.all.files)){
  colnames.all.files[i] <-
    sub("_", " ", sub("_", " ", colnames.all.files[i]))
}

#collects subfolder names for the sample names, input your pattern
colnames.merged.files <- list.files(main_folder)[(1:6)*3-2]
for (i in 1:length(colnames.merged.files)){
  colnames.merged.files[i] <-
    sub("_", " ", sub("_", " ", colnames.merged.files[i]))
}

```

Analysis for mRNA:

Collect two databases: one with individual replicates, one with merged replicates

```{r mRNA_all}

#makes a list of csv files:
arrays3D <- list.files(subfolder.names, pattern = "_160_10_Ex_Result",full.names = TRUE)

#reads the first file
first.file <- read.delim(arrays3D[1], header = TRUE, sep = ",", dec = ".")
colnames(first.file) <- c("cell", "1")

#cbinds the other files using first column to merge them
for(i in 2:(length(arrays3D))){
  new_file <- ""
  new_file <- read.delim(arrays3D[i], header = TRUE, sep = ",", dec = ".") 
  colnames(new_file) <- c("cell", (i))
  first.file <- merge(first.file, new_file, 
                    by = "cell", all = TRUE)
  
}

mRNA_all  <-  first.file[,-1]
colnames(mRNA_all) <- colnames.all.files

```

```{r mRNA_merged}

#makes a list of csv files:
arrays3D <- list.files(subfolder.names, pattern = "_160_10_Ex_Result",full.names = TRUE)

#reads the first 3 files
first.file <- read.delim(arrays3D[1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(first.file) <- c("cell", "1")

second.file <- read.delim(arrays3D[2], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(second.file) <- c("cell", "1")

third.file <- read.delim(arrays3D[3], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(third.file) <- c("cell", "1")

first.file <- rbind(first.file, second.file, third.file)
first.file[,1] <- 1:nrow(first.file)


#cbinds the other files using first column to merge them
for(i in 2:(length(arrays3D)/3)){
  
  file1 <- read.delim(arrays3D[3*i-2], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file1) <- c("cell", (3*i-2))
  
  file2 <- read.delim(arrays3D[3*i-1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file2) <- c("cell", (3*i-2))
  
  file3 <- read.delim(arrays3D[3*i], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file3) <- c("cell", (3*i-2))
  
  file1 <- rbind(file1, file2, file3)
  
  file1[,1] <- 1:nrow(file1)
  first.file <- merge(first.file, file1, 
                      by = "cell", all = TRUE)
  
}

mRNA_merged  <-  first.file[,-1]
colnames(mRNA_merged) <- colnames.merged.files
```

``` {r mRNA_over_0}

pct_positive <- sapply(mRNA_merged, function(x) {
  x <- x[!is.na(x)]          # keep only real cells
  100 * mean(x > 1)
})

result <- data.frame(
  condition   = names(pct_positive),
  n_cells     = sapply(mRNA_merged, function(x) sum(!is.na(x))),
  pct_mRNA_gt0 = round(pct_positive, 2),
  row.names = NULL
)
result

```

Plot:

```{r mRNA_all_plot}
#prepare for plotting
mRNA_all_gathered <- na.omit(gather(mRNA_all, key="sample",value = "mRNA_per_cell" ))
#plot
ggplot(mRNA_all_gathered, aes(y = mRNA_per_cell, x = sample)) +
  geom_violin(trim = TRUE, fill="peru", scale = "width",
             draw_quantiles = c(0.25, 0.5, 0.75)) +
  stat_summary_bin(fun = "mean", geom = "bar", colour = "black", fill = "white", alpha = 0.5)+
  labs(title="Distribution of mRNA",
       x ="", y = "mRNA per cell")+
  theme_pubr() +
  theme(axis.text.x=element_text(angle = 60, hjust = 1))
```

Figure S6C:

```{r mRNA_merged_plot}
#prepare for plotting
mRNA_merged_gathered <- na.omit(gather(mRNA_merged, key="sample",value = "mRNA_per_cell" ))
#plot
ggplot(mRNA_merged_gathered, aes(y = mRNA_per_cell, x = sample)) +
  geom_violin(trim = TRUE, fill="peru", scale = "width",
             draw_quantiles = c(0.25, 0.5, 0.75)) +
  stat_summary_bin(fun = "mean", geom = "bar", colour = "black", fill = "white", alpha = 0.5)+
  labs(title="Distribution of mRNA",
       x ="", y = "mRNA per cell")+
  theme_pubr() +
  theme(axis.text.x=element_text(angle = 60, hjust = 1))
```

```{r mRNA_for_Prism}

##for further Prism analysis:
mRNA_means <- data.frame(
  Column_Names = colnames(mRNA_all),
  Ave = sapply(mRNA_all, mean, na.rm = TRUE)
)
mRNA_means$Ave
```


Analysis for nascent transcription - counts number of transcription sites per cell:

Collect two databases: one with individual replicates, one with merged replicates

```{r TS_count_all}

#this is list of file names
arrays3D <- list.files(subfolder.names, pattern = "500_25_Vol_count_In_Result",full.names = TRUE)

first.file <- read.delim(arrays3D[1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(first.file) <- c("cell", "1")

#cbinds the other files using first column to merge them
for(i in 2:(length(arrays3D))){

  file1 <- read.delim(arrays3D[i], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file1) <- c("cell", (i))
  
  first.file <- merge(first.file, file1, 
                      by = "cell", all = TRUE)
} 

#remove first column, apply gather, omit NAs
TS_count.all <- first.file[,-1]
colnames(TS_count.all) <- colnames.all.files

TS_count.all_gathered <- na.omit(gather(TS_count.all, key = "cell", value = "TS_count"))

TS_count.all_gathered$ones <- 1
#order the data and some other data manipulations 
TS_count.all_gathered <- TS_count.all_gathered[order(TS_count.all_gathered$cell, TS_count.all_gathered$TS_count),]

#calculating frequency of specific numbers of TS 
options(dplyr.summarise.inform = FALSE)
TS_count.all_gathered <- TS_count.all_gathered %>% group_by(cell, TS_count) %>% summarise(Freq= n())

```

```{r TS_count_merged}

arrays3D <- list.files(subfolder.names, pattern = "500_25_Vol_count_In_Result",full.names = TRUE)

first.file <- read.delim(arrays3D[1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(first.file) <- c("cell", "1")

second.file <- read.delim(arrays3D[2], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(second.file) <- c("cell", "1")

third.file <- read.delim(arrays3D[3], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
colnames(third.file) <- c("cell", "1")

first.file <- rbind(first.file, second.file, third.file)

first.file[,1] <- 1:nrow(first.file)

#cbinds the other files using first column to merge them
for(i in 2:(length(arrays3D)/3)){
  file1 <- ""
  file2 <- ""
  file3 <- ""
  
  file1 <- read.delim(arrays3D[3*i-2], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file1) <- c("cell", (3*i-2))
  
  file2 <- read.delim(arrays3D[3*i-1], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file2) <- c("cell", (3*i-2))
  
  file3 <- read.delim(arrays3D[3*i], header = TRUE, sep = ",", dec = ".") # paste() merges strings for full path
  colnames(file3) <- c("cell", (3*i-2))
  
  file1 <- rbind(file1, file2, file3)
  
  file1[,1] <- 1:nrow(file1)
  first.file <- merge(first.file, file1, 
                      by = "cell", all = TRUE)
}

#remove first column, apply gather, omit NAs
TS_count.merged <- first.file[,-1]
colnames(TS_count.merged) <- colnames.merged.files

TS_count.merged_gathered <- na.omit(gather(TS_count.merged, key = "cell", value = "TS_count"))

TS_count.merged_gathered$ones <- 1
#order the data and some other data manipulations 
TS_count.merged_gathered <- TS_count.merged_gathered[order(TS_count.merged_gathered$cell, TS_count.merged_gathered$TS_count),]

#calculating frequency of specific numbers of TS 
options(dplyr.summarise.inform = FALSE)
TS_count.merged_gathered <- TS_count.merged_gathered %>% group_by(cell, TS_count) %>% summarise(Freq= n())

```


```{r TS_count_all_plot}
ggplot(TS_count.all_gathered, aes(fill = TS_count, y = Freq, x = cell))+
  scale_fill_gradient2(limits=c(0, 12), low=muted("dark red"), mid="floral white", high="light blue", midpoint=6)+
  geom_bar(position = "fill", stat = "identity")+
  labs(x ="", y = "Proportion of cells")+
  theme_pubr()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Figure S8D:

```{r TS_count_merged_plot}
ggplot(TS_count.merged_gathered, aes(fill = TS_count, y = Freq, x = cell))+
  scale_fill_gradient2(limits=c(0, 12), low=muted("dark red"), mid="floral white", high="light blue", midpoint=6)+
  geom_bar(position = "fill", stat = "identity")+
  labs(x ="", y = "Proportion of cells")+
  theme_pubr()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 

```


```{r TS_per_cell_for_Prism}

##for further Prism analysis:
TS_count_means <- data.frame(
  Column_Names = colnames(TS_count.all),
  Ave = sapply(TS_count.all, mean, na.rm = TRUE)
)
TS_count_means$Ave

```


