---
title: "M090_full_analysis_Fig3D"
author: "Anna Kogan"
date: "2026-05-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r libraries and wd, echo = FALSE, warning=FALSE,message=FALSE}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(tidyverse)
```

First: get the list of the results of the ImageJ analysis.
"Valid numbers" cleans up the data to exclude FOVs that didn't make it through the tracking 

```{r Lists of replicates}
path_d1_r1 = "E:/M090_new_analysis/M090_r1/M090v1_d1 tiff/analysis"
excels_d1_r1 <- list.files(path_d1_r1, pattern = ".csv", full.names = TRUE)
valid.numbers_r1 <- c(1:2, 5:8, 10:16, 19, 21:22, 31:38)
valid.files_d1_r1 <- excels_d1_r1[valid.numbers_r1]
path_d2_r1 = "E:/M090_new_analysis/M090_r1/M090v1_d2 tiff/analysis"
excels_d2_r1 <- list.files(path_d2_r1, pattern = ".csv", full.names = TRUE)
valid.files_d2_r1 <- excels_d2_r1[valid.numbers_r1]

path_d1_r2 = "E:/M090_new_analysis/M090_r4/M090v5_d1 tiff/analysis"
excels_d1_r2 <- list.files(path_d1_r2, pattern = ".csv", full.names = TRUE)
valid.numbers_r2 <- c(1:8,10,11,13:17,19,21:24,26,27,29:36,38:42,44,45)
valid.files_d1_r2 <- excels_d1_r2[valid.numbers_r2]
path_d2_r2 = "E:/M090_new_analysis/M090_r4/M090v5_d2 tiff/analysis"
excels_d2_r2 <- list.files(path_d2_r2, pattern = ".csv", full.names = TRUE)
valid.files_d2_r2 <- excels_d2_r2[valid.numbers_r2]

path_d1_r3 = "E:/M090_new_analysis/M090_r3/M090v4_d1 tiff/analysis"
excels_d1_r3 <- list.files(path_d1_r3, pattern = ".csv", full.names = TRUE)
valid.numbers_r3 <- c(1:3,7:9,11:12,14:17,19:26,30:32,34,36:40,42,44:46,48:49,51:55)
valid.numbers_r3 <- valid.numbers_r3[!valid.numbers_r3 %in% c(16,53,46,36,40)]
valid.files_d1_r3 <- excels_d1_r3[valid.numbers_r3]
path_d2_r3 = "E:/M090_new_analysis/M090_r3/M090v4_d2 tiff/analysis"
excels_d2_r3 <- list.files(path_d2_r3, pattern = ".csv", full.names = TRUE)
valid.files_d2_r3 <- excels_d2_r3[valid.numbers_r3]
```

Now read the data from the csv files in the folders to get the intensity values

``` {r Dataframes of replicates}
##### repl 1 processing #####
initial_intensities_r1 <- lapply(valid.files_d1_r1, function(file) {
  as.numeric(read.csv(file, header = TRUE)[,3])}) #reads mean intensities
progeny_intensities_r1 <- lapply(valid.files_d2_r1, function(file) {
  as.numeric(read.csv(file, header = TRUE)[,3])}) #reads mean intensities

numbers_rep_r1 <- rep(valid.numbers_r1, sapply(progeny_intensities_r1, length))
initial_rep_r1 <- rep(initial_intensities_r1, sapply(progeny_intensities_r1, length))
progeny_values_r1 <- unlist(progeny_intensities_r1)

# Combine into a data frame
plot_data_r1 <- data.frame(
  FOV_numbers = as.numeric(numbers_rep_r1),
  Initial_Intensity = as.numeric(initial_rep_r1),
  Progeny_Intensity = as.numeric(progeny_values_r1)
)


##### repl 2 processing #####
initial_intensities_r2 <- lapply(valid.files_d1_r2, function(file) {
  as.numeric(read.csv(file, header = TRUE)[,3])}) #reads mean intensities
progeny_intensities_r2 <- lapply(valid.files_d2_r2, function(file) {
  as.numeric(read.csv(file, header = TRUE)[,3])}) #reads mean intensities

numbers_rep_r2 <- rep(valid.numbers_r2, sapply(progeny_intensities_r2, length))
initial_rep_r2 <- rep(initial_intensities_r2, sapply(progeny_intensities_r2, length))
progeny_values_r2 <- unlist(progeny_intensities_r2)

# Combine into a data frame
plot_data_r2 <- data.frame(
  FOV_numbers = as.numeric(numbers_rep_r2),
  Initial_Intensity = as.numeric(initial_rep_r2),
  Progeny_Intensity = as.numeric(progeny_values_r2)
)


##### repl 3 processing #####
initial_intensities_r3 <- lapply(valid.files_d1_r3, function(file) {
  as.numeric(read.csv(file, header = TRUE)[,3])}) #reads mean intensities
progeny_intensities_r3 <- lapply(valid.files_d2_r3, function(file) {
  as.numeric(read.csv(file, header = TRUE)[,3])}) #reads mean intensities

numbers_rep_r3 <- rep(valid.numbers_r3, sapply(progeny_intensities_r3, length))
initial_rep_r3 <- rep(initial_intensities_r3, sapply(progeny_intensities_r3, length))
progeny_values_r3 <- unlist(progeny_intensities_r3)

# Combine into a data frame
plot_data_r3 <- data.frame(
  FOV_numbers = as.numeric(numbers_rep_r3),
  Initial_Intensity = as.numeric(initial_rep_r3),
  Progeny_Intensity = as.numeric(progeny_values_r3)
)
```

This chunk is optional and isn't used in the paper
Can be done to check the individual replicates


``` {r Optional non-normalised plots}

#making total distribution
initial_r1 <- data.frame(
  Y_pos = min(plot_data_r1$Progeny_Intensity)-100,
  Initial_Intensity = unique(plot_data_r1$Initial_Intensity)
)
progeny_r1 <- data.frame(
  X_pos = min(plot_data_r1$Initial_Intensity)-100,
  Progeny_Intensity = plot_data_r1$Progeny_Intensity
)

# Plot before normalisation
ggplot(data = plot_data_r1, aes(x = Initial_Intensity, y = Progeny_Intensity)) +
  #geom_boxplot(aes(group = FOV_numbers), fill = "lightgreen") + 
  geom_point(fill = "black") +
  geom_line(data = plot_data_r1, aes(x = Initial_Intensity, y = Progeny_Intensity, group = FOV_numbers))+
  geom_boxplot(data = initial_r1, aes(y=Y_pos, x = Initial_Intensity), width = 100, outliers = F)+
  geom_boxplot(data = progeny_r1, aes(x=X_pos, y = Progeny_Intensity), width = 100, outliers = F)+
  theme_pubr()

##check that these are the same: 
#length(unlist(initial_intensities_r2))
#length(unique(plot_data_r2$Initial_Intensity))

#making total distribution
initial_r2 <- data.frame(
  Y_pos = min(plot_data_r2$Progeny_Intensity)-100,
  Initial_Intensity = unique(plot_data_r2$Initial_Intensity)
)
progeny_r2 <- data.frame(
  X_pos = min(plot_data_r2$Initial_Intensity)-100,
  Progeny_Intensity = plot_data_r2$Progeny_Intensity
)

# Plot before normalisation
ggplot(data = plot_data_r2, aes(x = Initial_Intensity, y = Progeny_Intensity)) +
  geom_boxplot(aes(group = FOV_numbers), fill = "lightgreen") + 
  geom_point(fill = "black") +
  geom_boxplot(data = initial_r2, aes(y=Y_pos, x = Initial_Intensity), width = 100, outliers = F)+
  geom_boxplot(data = progeny_r2, aes(x=X_pos, y = Progeny_Intensity), width = 100, outliers = F)+
  theme_pubr()


##check that these are the same: 
#length(unlist(initial_intensities_r3))
#length(unique(plot_data_r3$Initial_Intensity))

#making total distribution
initial_r3 <- data.frame(
  Y_pos = min(plot_data_r3$Progeny_Intensity)-100,
  Initial_Intensity = unique(plot_data_r3$Initial_Intensity)
)
progeny_r3 <- data.frame(
  X_pos = min(plot_data_r3$Initial_Intensity)-50,
  Progeny_Intensity = plot_data_r3$Progeny_Intensity
)

# Plot before normalisation
ggplot(data = plot_data_r3, aes(x = Initial_Intensity, y = Progeny_Intensity)) +
  geom_boxplot(aes(group = FOV_numbers), fill = "lightgreen") + 
  geom_point(fill = "black") +
  geom_boxplot(data = initial_r3, aes(y=Y_pos, x = Initial_Intensity), width = 100, outliers = F)+
  geom_boxplot(data = progeny_r3, aes(x=X_pos, y = Progeny_Intensity), width = 30, outliers = F)+
  theme_pubr()
```

Because the average intensity differs between the replicates,
we normalise these based on the mean and the sd. 

``` {r Normalisation}

subtr_d1_r1 <- mean(plot_data_r1$Initial_Intensity)
div_d1_r1 <- sd(plot_data_r1$Initial_Intensity)
subtr_d2_r1 <- mean(plot_data_r1$Progeny_Intensity)
div_d2_r1 <- sd(plot_data_r1$Progeny_Intensity)

plot_data_sd_r1 <- data.frame(
  FOV_numbers = plot_data_r1$FOV_numbers,
  Initial_Intensity = (plot_data_r1$Initial_Intensity - subtr_d1_r1)/div_d1_r1*400,
  Progeny_Intensity = (plot_data_r1$Progeny_Intensity - subtr_d2_r1)/div_d2_r1*400
)

subtr_d1_r2 <- mean(plot_data_r2$Initial_Intensity)
div_d1_r2 <- sd(plot_data_r2$Initial_Intensity)
subtr_d2_r2 <- mean(plot_data_r2$Progeny_Intensity)
div_d2_r2 <- sd(plot_data_r2$Progeny_Intensity)

plot_data_sd_r2 <- data.frame(
  FOV_numbers = plot_data_r2$FOV_numbers,
  Initial_Intensity = (plot_data_r2$Initial_Intensity - subtr_d1_r2)/div_d1_r2*400,
  Progeny_Intensity = (plot_data_r2$Progeny_Intensity - subtr_d2_r2)/div_d2_r2*400
)

subtr_d1_r3 <- mean(plot_data_r3$Initial_Intensity)
div_d1_r3 <- sd(plot_data_r3$Initial_Intensity)
subtr_d2_r3 <- mean(plot_data_r3$Progeny_Intensity)
div_d2_r3 <- sd(plot_data_r3$Progeny_Intensity)

plot_data_sd_r3 <- data.frame(
  FOV_numbers = plot_data_r3$FOV_numbers,
  Initial_Intensity = (plot_data_r3$Initial_Intensity - subtr_d1_r3)/div_d1_r3*400,
  Progeny_Intensity = (plot_data_r3$Progeny_Intensity - subtr_d2_r3)/div_d2_r3*400
)
```


Now to combine all the replicates into a single dataframe: 

``` {r Single dataframe}
repl1 <- plot_data_sd_r1
repl1$repl <- 1
repl2 <- plot_data_sd_r2
repl2$repl <- 2
repl3 <- plot_data_sd_r3
repl3$repl <- 3

repl2$FOV_numbers <- repl2$FOV_numbers + max(repl1$FOV_numbers)
repl3$FOV_numbers <- repl3$FOV_numbers + max(repl2$FOV_numbers)

total_data <- rbind(repl1, repl2, repl3)

##for the barplots of total distribution: 
initial <- data.frame(
  Y_pos = min(total_data$Progeny_Intensity)-50,
  Initial_Intensity = unique(total_data$Initial_Intensity)
)
progeny <- data.frame(
  X_pos = min(total_data$Initial_Intensity)-50,
  Progeny_Intensity = total_data$Progeny_Intensity
)

```


``` {r Plotting the normalised replicates}

#### normalised individual dots per replicate
ggplot(data = total_data[(total_data$repl == 1),], aes(x = round(Initial_Intensity), 
                                                       y = round(Progeny_Intensity))) +
  geom_point(aes(color = as.factor(repl)),size = 1.5) +
  geom_smooth(colour="black")+#(aes(color = as.factor(repl)))+
  stat_cor(aes(color = as.factor(repl)),method="pearson")+
  geom_line(data = total_data[(total_data$repl == 1),], aes(x = round(Initial_Intensity), 
                                                            y = round(Progeny_Intensity), group = FOV_numbers))+
  #geom_boxplot(data = initial, aes(y=Y_pos, x = Initial_Intensity), width = 50, outliers = F)+
  #geom_boxplot(data = progeny, aes(x=X_pos, y = Progeny_Intensity), width = 20, outliers = F)+
  #geom_label(aes(label = FOV_numbers), position = "nudge") +
theme_pubr()

ggplot(data = total_data[(total_data$repl == 2),], aes(x = round(Initial_Intensity), 
                                                       y = round(Progeny_Intensity))) +
  geom_point(aes(color = as.factor(repl)),size = 1.5) +
  geom_smooth(colour="black")+#(aes(color = as.factor(repl)))+
  stat_cor(aes(color = as.factor(repl)),method="pearson")+
  geom_line(data = total_data[(total_data$repl == 2),], aes(x = round(Initial_Intensity), 
                                                            y = round(Progeny_Intensity), group = FOV_numbers))+
  #geom_boxplot(data = initial, aes(y=Y_pos, x = Initial_Intensity), width = 50, outliers = F)+
  #geom_boxplot(data = progeny, aes(x=X_pos, y = Progeny_Intensity), width = 20, outliers = F)+
  #geom_label(aes(label = FOV_numbers), position = "nudge") +
  theme_pubr()

ggplot(data = total_data[(total_data$repl == 3),], aes(x = round(Initial_Intensity), 
                                                       y = round(Progeny_Intensity))) +
  geom_point(aes(color = as.factor(repl)),size = 1.5) +
  geom_smooth(colour="black")+#(aes(color = as.factor(repl)))+
  stat_cor(aes(color = as.factor(repl)),method="pearson")+
  geom_line(data = total_data[(total_data$repl == 3),], aes(x = round(Initial_Intensity), 
                                                            y = round(Progeny_Intensity), group = FOV_numbers))+
  #geom_boxplot(data = initial, aes(y=Y_pos, x = Initial_Intensity), width = 50, outliers = F)+
  #geom_boxplot(data = progeny, aes(x=X_pos, y = Progeny_Intensity), width = 20, outliers = F)+
  #geom_label(aes(label = FOV_numbers), position = "nudge") +
  theme_pubr()
```


Creating the main plot for the paper

``` {r Plotting the total data}

total_data_summ <- total_data %>% group_by(FOV_numbers) %>% summarise(
  Initial_Intensity = mean(Initial_Intensity),
  prog_mean = mean(Progeny_Intensity),
  prog_sd = sd(Progeny_Intensity),
  repl = mean(repl)
  )

ggplot(data = total_data_summ, aes(x = round(Initial_Intensity), y = round(prog_mean))) +
  geom_point(aes(color = as.factor(repl)),size = 1.5) +
  geom_smooth(fill = "white", colour = "black") +
  stat_cor(aes(color = as.factor(repl)),method="pearson")+
  geom_errorbar(aes(ymin = (prog_mean - prog_sd),ymax = (prog_mean + prog_sd)), width = 10) +
  theme_pubr()


ggplot() +
  geom_point(data = total_data_summ, aes(x = round(Initial_Intensity), y = round(prog_mean), 
                                         color = as.factor(repl)),size = 1.5) +
  geom_smooth(data = total_data, aes(x = round(Initial_Intensity), y = round(Progeny_Intensity)), 
              fill = "white", colour="black")+
  stat_cor(data = total_data, aes(x = round(Initial_Intensity), y = round(Progeny_Intensity)),method="pearson")+
  geom_errorbar(data = total_data_summ, aes(x = round(Initial_Intensity), ymin = (prog_mean - prog_sd),
                                            ymax = (prog_mean + prog_sd)), width = 40) +
  xlab("Founder Intensity, A.U.")+
  ylab("Progeny Intenisty, A.U.")+
  scale_x_continuous(breaks=c(-500,0,500, 1000), limits = c(-900,2000))+
  scale_y_continuous(breaks=c(-500,0,500,1000,1500,2000),limits = c(-750,1800)) +
  theme_pubr()

```





``` {r Standard deviation plots}

total_data_sd <- total_data %>% group_by(FOV_numbers) %>% summarise(
  Initial_Intensity = mean(Initial_Intensity),
  prog_sd = sd(Progeny_Intensity),
  repl = mean(repl)
)

ggplot(data = total_data_sd, aes(x = round(Initial_Intensity), y = round(prog_sd))) +
  geom_point(aes(color = as.factor(repl)),size = 1.5) +
  stat_cor(aes(color = as.factor(repl)),method="pearson")+
  theme_pubr()

ggplot(data = total_data_sd, aes(x = round(Initial_Intensity), y = round(prog_sd))) +
  geom_point(aes(color = as.factor(repl)),size = 1.5) +
  stat_cor(method="pearson")+
  theme_pubr()

```

``` {r Number of progeny plots}

total_data_N <- total_data %>% group_by(FOV_numbers) %>% summarise(
  Initial_Intensity = mean(Initial_Intensity),
  prog_hrs = 96/log2(n()),
  repl = mean(repl)
)

ggplot(data = total_data_N, aes(x = round(Initial_Intensity), y = prog_hrs)) +
  geom_point(aes(color = as.factor(repl)),size = 1.5) +
  stat_cor(aes(color = as.factor(repl)),method="pearson")+
  scale_y_continuous(breaks=c(24, 48),limits = c(15,60)) +
  theme_pubr()

ggplot(data = total_data_N, aes(x = round(Initial_Intensity), y = prog_hrs)) +
  geom_point(aes(color = as.factor(repl)),size = 1.5) +
  stat_cor(method="pearson")+
  scale_y_continuous(breaks=c(24, 48),limits = c(15,60)) +
  theme_pubr()

```
``` {r from Claude for division times calculations}

total_data_N <- total_data %>% group_by(FOV_numbers) %>% summarise(
  Initial_Intensity = mean(Initial_Intensity),
  prog_N = n(),
  repl = mean(repl)
)

# 1) Growth rate per FOV (divisions/hour), assuming exponential growth, no death
fov <- total_data_N %>%
  mutate(rate = log2(prog_N) / 96)

# 2) Average growth rate per replicate, then invert to get division time
per_repl <- fov %>%
  group_by(repl) %>%
  summarise(
    n_FOV    = n(),
    rate     = mean(rate),
    div_time = 1 / rate,                     # hours per division
    div_3d   = 72  / div_time,               # divisions in a 3-day window
    div_6d   = 144 / div_time,               # divisions in a 6-day window
    .groups  = "drop"
  )
print(per_repl)

# 3) Mean +- SD across replicates
summ <- per_repl %>%
  summarise(across(c(div_time, div_3d, div_6d),
                   list(mean = mean, sd = sd)))
print(summ)

# 4) Fill in the sentence
with(summ, cat(sprintf(
  paste0("Furthermore, from the number of progeny, we derived an average cell division ",
         "time of %.1f \u00b1 %.1f hours, which corresponds to %.1f \u00b1 %.1f cell divisions ",
         "in a 3-day memory window or %.1f \u00b1 %.1f in a standard 6-day memory window, ",
         "assuming no cell death.\n"),
  div_time_mean, div_time_sd, div_3d_mean, div_3d_sd, div_6d_mean, div_6d_sd)))

```


